---
title: "Hawaii's Big Island"
subtitle: "Key Insights and Trends"
author: "Phoenix Stout"
date: "December 15, 2025"
format:
  revealjs:
    theme: serif
    slide-number: true
    preview-links: auto
    logo: ""
    css: custom.css
    footer: "MA615 Final Project | Phoenix Stout"
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

```{r setup}
library(tidyverse)
library(knitr)
library(scales)
library(plotly)

# Source helper functions
source("../scripts/02_data_processing.R")

# Load all data
data <- load_all_data()

# Color palette
hawaii_colors <- c(
  ocean = "#006BA6",
  lava = "#C1121F",
  forest = "#2D6A4F",
  sand = "#F4A261",
  sunset = "#E76F51",
  sky = "#219EBC"
)
```

## Overview {background-color="#006BA6"}

::::: columns
::: {.column width="50%"}
### Hawaii's Big Island

-   Largest Hawaiian Island
-   4,028 square miles
-   206,400 residents (2023)
-   Active volcanic zones
-   Diverse microclimates
:::

::: {.column width="50%"}
### Analysis Focus

1.  Demographics
2.  Climate
3.  Economics
4.  Natural Disasters
:::
:::::

## Executive Summary

::: incremental
-   **Population**: Growing 4.9% since 2015, but aging rapidly
-   **Economy**: Tourism-dependent with \$70,200 median income
-   **Climate**: Warming trends, extreme heat days tripled
-   **Hazards**: \$800M in volcanic damage (2018), ongoing tsunami risk
:::

# Demographics {background-color="#2D6A4F"}

## Population Growth

```{r, warning=FALSE}
#| fig-width: 10
#| fig-height: 6

p <- ggplot(data$demographics, aes(x = year, y = population)) +
  geom_line(color = hawaii_colors["ocean"], size = 2) +
  geom_point(color = hawaii_colors["ocean"], size = 4) +
  scale_y_continuous(labels = comma, limits = c(190000, 210000)) +
  labs(
    title = "Steady Population Growth",
    subtitle = "196,823 (2015) → 206,400 (2023)",
    x = NULL,
    y = "Population"
  ) +
  theme_minimal(base_size = 16) +
  theme(plot.title = element_text(face = "bold", size = 20))

ggplotly(p) %>%
  layout(hovermode = "x unified")
```

## Aging Population

::::: columns
::: {.column width="50%"}
```{r}
#| fig-width: 6
#| fig-height: 5

aging_data <- data$demographics %>%
  select(year, percent_65_over, percent_under_18) %>%
  pivot_longer(cols = c(percent_65_over, percent_under_18),
               names_to = "group", values_to = "percentage") %>%
  mutate(group = ifelse(group == "percent_65_over", "65 and Over", "Under 18"))

ggplot(aging_data, aes(x = year, y = percentage, color = group)) +
  geom_line(size = 1.5) +
  geom_point(size = 3) +
  scale_color_manual(values = c("65 and Over" = "#8B0000",
                                "Under 18" = "#1E90FF")) +
  labs(
    title = "Age Distribution Shift",
    x = NULL,
    y = "Percentage",
    color = "Age Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```

:::

::: {.column width="50%" style="font-size: 0.6em;"}
**Key Trends**

-   Median age: 38.2 → 40.5
-   Seniors (65+): 17.2% → 21.1%
-   Youth (<18): 21.5% → 19.3%

**Implications**

-   Healthcare demands ↑
-   School enrollment ↓
-   Workforce challenges
:::
:::::

## Cultural Diversity

```{r}
#| fig-width: 10
#| fig-height: 6

ggplot(data$race_composition, aes(x = reorder(category, percentage), y = percentage)) +
  geom_col(fill = hawaii_colors["sunset"], width = 0.7) +
  geom_text(aes(label = paste0(percentage, "%")),
            hjust = -0.2, size = 6, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(limits = c(0, 40), expand = c(0, 0)) +
  labs(
    title = "Population Distribution by Race",
    x = NULL,
    y = "Percentage of Population"
  ) +
  theme_minimal(base_size = 16) +
  theme(plot.title = element_text(face = "bold", size = 18))
```

# Climate {background-color="#219EBC"}

## Monthly Patterns

```{r}
#| fig-width: 11
#| fig-height: 6

climate_data <- data$climate_monthly %>%
  mutate(month = factor(month, levels = month.name))

ggplot(climate_data) +
  geom_col(aes(x = month, y = avg_precip_in),
           fill = hawaii_colors["ocean"], alpha = 0.6, width = 0.7) +
  geom_line(aes(x = month, y = avg_temp_f / 5, group = 1),
            color = hawaii_colors["lava"], size = 1.5) +
  geom_point(aes(x = month, y = avg_temp_f / 5),
             color = hawaii_colors["lava"], size = 4) +
  scale_y_continuous(
    name = "Precipitation (inches)",
    sec.axis = sec_axis(~ . * 5, name = "Temperature (°F)")
  ) +
  labs(
    title = "Hilo Climate: Stable Temperature, Variable Rainfall",
    x = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    plot.title = element_text(face = "bold", size = 17),
    axis.title.y.left = element_text(color = hawaii_colors["ocean"], size = 14),
    axis.title.y.right = element_text(color = hawaii_colors["lava"], size = 14)
  )
```

## Climate Change Signals

```{r}
#| fig-width: 11
#| fig-height: 6

temp_min <- 70
temp_max <- 95

# Scale extreme heat days to fit in the temperature range
heat_max <- max(data$climate_annual$extreme_heat_days, na.rm = TRUE)
scale_factor <- (temp_max - temp_min) / heat_max

ggplot(data$climate_annual, aes(x = year)) +
  # Use geom_rect instead of geom_col to control where bars start
  geom_rect(aes(xmin = year - 0.4, xmax = year + 0.4,
                ymin = temp_min,
                ymax = temp_min + extreme_heat_days * scale_factor),
            fill = hawaii_colors["sunset"], alpha = 0.5) +
  geom_line(aes(y = avg_temp_f), color = hawaii_colors["lava"], size = 1.8) +
  geom_point(aes(y = avg_temp_f), color = hawaii_colors["lava"], size = 4) +
  scale_y_continuous(
    name = "Average Temperature (°F)",
    limits = c(temp_min, temp_max),
    sec.axis = sec_axis(~ (. - temp_min) / scale_factor, name = "Extreme Heat Days")
  ) +
  labs(
    title = "Warming Trend: Extreme Heat Days Tripled",
    subtitle = "12 days (2015) → 38 days (2023)",
    x = "Year"
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold", size = 18))
```

::: notes
Temperature increased 2.3°F from 2015 to 2023. Extreme heat days (above
90°F) more than tripled.
:::

# Economics {background-color="#F4A261"}

## Income Growth

```{r}
#| fig-width: 10
#| fig-height: 6

p2 <- ggplot(data$economics, aes(x = year, y = median_income)) +
  geom_line(color = hawaii_colors["forest"], size = 2) +
  geom_point(color = hawaii_colors["forest"], size = 4) +
  scale_y_continuous(labels = dollar, limits = c(50000, 75000)) +
  labs(
    title = "Median Income: 29.5% Growth",
    subtitle = "$54,200 → $70,200",
    x = NULL,
    y = "Median Household Income"
  ) +
  theme_minimal(base_size = 16) +
  theme(plot.title = element_text(face = "bold", size = 20))

ggplotly(p2) %>%
  layout(hovermode = "x unified")
```

## Unemployment Recovery

```{r}
#| fig-width: 10
#| fig-height: 6

ggplot(data$economics, aes(x = year, y = unemployment_rate)) +
  geom_area(fill = hawaii_colors["lava"], alpha = 0.3) +
  geom_line(color = hawaii_colors["lava"], size = 2) +
  geom_point(color = hawaii_colors["lava"], size = 4) +
  geom_hline(yintercept = 4.0, linetype = "dashed", color = "gray30", size = 1) +
  annotate("rect", xmin = 2019.5, xmax = 2020.5, ymin = 0, ymax = 10,
           alpha = 0.15, fill = "red") +
  annotate("text", x = 2020, y = 9.5, label = "COVID-19",
           size = 6, fontface = "bold") +
  scale_y_continuous(labels = percent_format(scale = 1), limits = c(0, 10)) +
  labs(
    title = "Unemployment: COVID Spike, Strong Recovery",
    x = "Year",
    y = "Unemployment Rate (%)"
  ) +
  theme_minimal(base_size = 16) +
  theme(plot.title = element_text(face = "bold", size = 19))
```

## Industry Breakdown

::::: columns
::: {.column width="55%"}
```{r}
#| fig-width: 6
#| fig-height: 4.5

industry_plot <- data$industry_composition %>%
  arrange(desc(employment_share)) %>%
  slice(1:5)

ggplot(industry_plot, aes(x = reorder(sector, employment_share),
                          y = employment_share)) +
  geom_col(fill = hawaii_colors["ocean"], width = 0.7) +
  geom_text(aes(label = paste0(employment_share, "%")),
            hjust = -0.2, size = 5, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(limits = c(0, 35)) +
  labs(
    title = "Top 5 Economic Sectors",
    x = NULL,
    y = "Employment Share (%)"
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))
```

:::

::: {.column width="45%"}
**Tourism Dominance**

-   28.5% of jobs
-   Avg wage: \$42,000
-   Shock vulnerability

**Higher-Paying Sectors**

-   Healthcare: \$62,000
-   Government: \$58,000
-   Construction: \$52,000
:::
:::::

# Natural Disasters {background-color="#C1121F"}

## Disaster Impact

```{r, warning=FALSE}
#| fig-width: 11
#| fig-height: 6

ggplot(data$disasters, aes(x = year, y = damage_millions, fill = type)) +
  geom_col() +
  scale_y_continuous(labels = dollar_format(suffix = "M"), limits = c(0, 850)) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Economic Impact of Natural Disasters",
    subtitle = "2018 Kilauea Eruption: $800M damage, 2,500 evacuations",
    x = "Year",
    y = "Damage (Millions USD)",
    fill = "Disaster Type"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 18)
  )
```

## Volcanic Activity

```{r}
#| fig-width: 10
#| fig-height: 6

ggplot(data$volcanic_activity, aes(x = year)) +
  geom_col(aes(y = eruption_days, fill = "Eruption Days"), alpha = 0.7) +
  geom_line(aes(y = earthquakes_m3plus / 5, color = "Earthquakes (M3+)"), size = 1.5) +
  geom_point(aes(y = earthquakes_m3plus / 5, color = "Earthquakes (M3+)"), size = 4) +
  scale_fill_manual(name = NULL, values = c("Eruption Days" = unname(hawaii_colors["lava"]))) +
  scale_color_manual(name = NULL, values = c("Earthquakes (M3+)" = unname(hawaii_colors["ocean"]))) +
  scale_y_continuous(
    name = "Eruption Days per Year",
    sec.axis = sec_axis(~ . * 5, name = "Earthquakes (M3+)")
  ) +
  labs(
    title = "Kilauea Activity: Eruptions and Seismic Events",
    x = "Year"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 17),
    legend.position = "bottom"
  )
```

::: notes
2018 saw peak activity with 214 eruption days and 2,345 earthquakes M3+
:::

## Tsunami Risk

::::: columns
::: {.column width="50%" style="font-size: 0.8em;"}
### Population at Risk

| Zone                | Population |
|---------------------|------------|
| **Evacuation Zone** | **12,000** |
| High Risk           | 35,000     |
| Medium Risk         | 58,000     |
| Low Risk            | 101,400    |

**Total High/Evac: 47,000** (23% of population)
:::

::: {.column width="50%" style="font-size: 0.8em;"}
### Risk Factors

-   Pacific Ring of Fire
-   Low-lying coastal areas
-   Limited evacuation routes
-   24/7 warning system active

### Recent Events

-   2011: Japan tsunami
-   Regular warning drills
-   Updated evacuation maps
:::
:::::

# Key Findings {background-color="#2D6A4F"}

## Synthesis

::: incremental
**Demographic Challenge** - Aging population (21% seniors) → healthcare
pressures - Declining youth population → workforce concerns

**Economic Vulnerability** - Tourism dependence (28.5%) → shock
sensitivity - Strong recovery post-COVID → resilience

**Climate Concerns** - Clear warming trend (+2.3°F) - Extreme heat days
tripled → adaptation needed

**Persistent Hazards** - Active volcanoes (\$800M+ damages) - 47,000 in
tsunami risk zones
:::

## Strategic Priorities

::::: columns
::: {.column width="50%"}
### Short Term

1.  Disaster preparedness
2.  Climate adaptation
3.  Tourism recovery support
4.  Healthcare capacity
:::

::: {.column width="50%"}
### Long Term

1.  Economic diversification
2.  Sustainable development
3.  Infrastructure resilience
4.  Demographic planning
:::
:::::

## Conclusions {background-color="#006BA6"}


::::: columns
::: {.column width="50%"}
### Strengths

-   Economic recovery
-   Cultural diversity
-   Natural beauty
-   Tourism appeal
:::

::: {.column width="50%"}
### Challenges

-   Climate change
-   Natural hazards
-   Aging population
-   Economic concentration
:::
:::::


::: center
**Phoenix Stout**

MA615 Final Project

December 15, 2025

[GitHub Repository](https://github.com/stoutphoenix/Hawaii-BigIsland)
:::
